Tutorial_2a

### 2024-11-24 Tutorial_2a.Rmd
update tut 2a

Tutorial 2a shows various ways to fit Bayesian binary regression models to discrete time-to-event data. The data are taken from the first experiment of Panis and Schmidt (2016), but using only the no-mask trials (prime = blank, congruent, or incongruent). We then compare models using WAIC and LOO, and interpret parameter estimates for one model. 
We also plot the logit and cloglog link functions, and illustrate how various prior distributions on the logit and cloglog scales look on the original probability (i.e., hazard) scale.

@@

```{r data-priors-M0}
M0_data <- ptb_data %>% select(pid, event, timebin)

priors_M0 <- c(
  set_prior("skew_normal(-0.59,1.26,-4.22)", class = "b"), # See suppl. material 
  set_prior("normal(0, 1)", class = "sd"),                    
  set_prior("lkj(2)", class = "cor")                        
)
```

@@

Fit model M0.

```{r fit-model-M0, eval=F} 
plan(multicore)

model_M0 <-                     # wrong data file name
   brm(data = M0_data,
       family = binomial(link="cloglog"),
       event|trials(1) ~ 0 + timebin + (0 + timebin | pid),
       prior = priors_M0,
       chains = 4, cores = 4, iter = 3000, warmup = 1000,
       control = list(adapt_delta = 0.999, step_size = 0.04, max_treedepth = 12),
       seed = 12, init = "0",
       file = "Tutorial_2_Bayesian/models/model_M0")
```

This took 43 minutes on a MacBook Pro (Sonoma 14.6.1 OS, 18GB Memory, M3 Pro Chip).

```{r}
model_M0 <- readRDS("Tutorial_2_Bayesian/models/model_M0.rds")
summary(model_M0)
formula(model_M0)
plot(model_M0)
```

@@

```{r data-priors-M0d}
M0d_data <- ptb_data %>% select(pid, event, d6:d8, d10:d15)

@@

model_M0d <-            # bernouilli.  :(
   brm(data = M0d_data,
       family = bernoulli(link="cloglog"),
       event ~ 0 + d6 + d7 + d8 + Intercept + d10 + d11 + d12 + d13 + d14 + d15 + 
                   (d6 + d7 + d8 + 1 + d10 + d11 + d12 + d13 + d14 + d15  | pid),
       prior = priors_M0d,
       chains = 4, cores = 4, iter = 3000, warmup = 1000,
       control = list(adapt_delta = 0.999, step_size = 0.04, max_treedepth = 12),
       seed = 12, init = "0",
       file = "Tutorial_2_Bayesian/models/model_M0d")
```

M0d took 77 minutes.

```{r}
model_M0d <- readRDS("Tutorial_2_Bayesian/models/model_M0d.rds")
summary(model_M0d)
formula(model_M0d)
```

@@

```{r data-priors-M0c}
M0c_data <- ptb_data %>% select(pid, event, period_9) %>%
  mutate(period_9_sq = period_9^2,
         period_9_cu = period_9^3)

priors_M0c <- c(
  set_prior("skew_normal(-0.2,0.71,-2.2)", class = "b"),      
  set_prior("skew_normal(-1,1,-2)", class = "b", coef = "Intercept"),  
  set_prior("normal(0, 1)", class = "sd"),  
  set_prior("lkj(2)", class = "cor")  
)
```

@@

model_M0c <-
   brm(data = M0c_data,
       family = binomial(link="cloglog"),
       event | trials(1) ~ 0 + Intercept + period_9 + period_9_sq + period_9_cu + 
                          (0 + Intercept + period_9 + period_9_sq + period_9_cu  | pid),
       prior = priors_M0c,
       chains = 4, cores = 4, iter = 3000, warmup = 1000,
       control = list(adapt_delta = 0.999, step_size = 0.04, max_treedepth = 12),
       seed = 12, init = "0",
       file = "Tutorial_2_Bayesian/models/model_M0c")
```

Model_M0c took about 144 minutes to run.

@@

```{r}
dat <- tibble(x = -3:6,
              x2 = x^2,
              x3 = x^3,
              cloglog1 = -1.8   + 0.3*x   - 0.1*x2   + 0.009*x3,
              cloglog2 = -1     + 0.5*x   - 0.1*x2   + 0.004*x3,
              cloglog3 = -3     - 0.1*x   + 0.12*x2  - 0.002*x3,
              cloglog4 = -0.40  + 0.2*x   - 0.04*x2  + 0.003*x3,
              haz1 = 1 - exp(-1*exp(cloglog1)),
              haz2 = 1 - exp(-1*exp(cloglog2)),
              haz3 = 1 - exp(-1*exp(cloglog3)),
              haz4 = 1 - exp(-1*exp(cloglog4)))

@@

set_prior("normal(0.1,.08)",class="b",coef="period_9"), # ok !!!
set_prior("normal(0,.06)",class="b",coef="period_9_sq"),

@@

model_M0_lin_prior448 <-
   brm(data = M0c_data,
       family = binomial(link="cloglog"),
       event | trials(1) ~ 0 + Intercept + period_9 + period_9_sq +
                          (0 + Intercept + period_9 + period_9_sq | pid),
       prior = priors_M0c_prior,
       chains = 4, cores = 4, iter = 3000, warmup = 1000,
       control = list(adapt_delta = 0.999, step_size = 0.04, max_treedepth = 12),
       seed = 12, init = "0",
       sample_prior = "only",
       file = "Tutorial_2_Bayesian/models/model_M0_lin_prior448")
```

```{r}
post <- as_draws_df(model_M0_lin_prior448) %>% 
  select(-lp__) %>% 
  as_tibble()

@@

# plot
p_tidy_fixed <- ggplot(tidy_fixed, aes(x = value, y = name)) +  
  stat_halfeye(alpha=0.7,point_interval = "median_qi",.width=c(.8,.99) ) 
p_tidy_fixed

@@

N_obs = 150 # increase for smoother results

# prepare columns of data set
df_pred <- tibble(period_9 = integer(0),
                 period_9_sq = integer(0),
                 pid = integer(0),
                  .row = integer(0),
                  .chain = integer(0),
                  .iteration = integer(0),
                  .draw = integer(0),
                  .prediction = integer(0),
                  obs = integer(0))

@@

for(i in 1:N_obs){
  prior_pred <- add_predicted_draws(model_M0_lin_prior448, 
                                    newdata = newdata_prior, 
                                    summary=F, 
                                    ndraws=NULL) %>% 
                mutate(obs = i)
  df_pred <- bind_rows(df_pred, prior_pred)
}

@@

ggplot(df_pred_haz %>% 
   filter(.draw <7639), aes(x=period_9, y=pred_haz, group=.draw)) +
   geom_line(color="black") +
   geom_line(data=df_pred_haz %>% filter(.draw ==7418), 
             color="red", linewidth=2) +
  geom_line(data=df_pred_haz %>% filter(.draw == 6001), 
            color="green", linewidth=2) +
  geom_line(data=df_pred_haz %>% filter(.draw == sel[3]), 
          color="blue", linewidth=2) +
  scale_y_continuous(limits = c(0,1)) +
  scale_x_continuous(limits = c(-3,6), breaks = c(-3:6)) +
  facet_wrap(~pid)
```

>>>>

Tutorial 2a shows various ways to fit Bayesian binary regression models to discrete time-to-event data. The data are taken from the first experiment of Panis and Schmidt (2016), but using only the no-mask trials (prime = blank, congruent, or incongruent). We perform prior predictive checks to infer the prior distributions we will employ. We compare models using WAIC and LOO, and interpret parameter estimates for one model. 
We end by plotting the logit and cloglog link functions, and illustrate how various prior distributions on the logit and cloglog scales look on the original probability (i.e., hazard) scale.

@@

```{r data-priors-M0}
data_M0i <- ptb_data %>% select(pid, event, timebin)

priors_M0i <- c(
  set_prior("skew_normal(-1,1,-2)", class = "b"), 
  set_prior("normal(0, 1)", class = "sd"),                    
  set_prior("lkj(2)", class = "cor")                        
)
```

@@

Fit model M0i.

```{r fit-model-M0i, eval=F} 
plan(multicore)

model_M0i <-                    
   brm(data = data_M0i,
       family = bernoulli(link="cloglog"),
       formula = event ~ 0 + timebin + (0 + timebin | pid),
       prior = priors_M0i,
       chains = 4, cores = 4, iter = 3000, warmup = 1000,
       control = list(adapt_delta = 0.999, step_size = 0.04, max_treedepth = 12),
       seed = 12, init = "0",
       file = "Tutorial_2_Bayesian/models/model_M0i")
```

This took 28 minutes on a MacBook Pro (Sonoma 14.6.1 OS, 18GB Memory, M3 Pro Chip).

```{r}
model_M0i <- readRDS("Tutorial_2_Bayesian/models/model_M0i.rds")
summary(model_M0i)
formula(model_M0i)
plot(model_M0i)
```

@@

```{r data-priors-M0d}
data_M0d <- ptb_data %>% select(pid, event, d6:d8, d10:d15)

@@

model_M0d <-      
   brm(data = data_M0d,
       family = bernoulli(link="cloglog"),
       event ~ 0 + d6 + d7 + d8 + Intercept + d10 + d11 + d12 + d13 + d14 + d15 + 
                   (d6 + d7 + d8 + 1 + d10 + d11 + d12 + d13 + d14 + d15  | pid),
       prior = priors_M0d,
       chains = 4, cores = 4, iter = 3000, warmup = 1000,
       control = list(adapt_delta = 0.999, step_size = 0.04, max_treedepth = 12),
       seed = 12, init = "0",
       file = "Tutorial_2_Bayesian/models/model_M0d")
```

M0d took about 77 minutes.

```{r}
model_M0d <- readRDS("Tutorial_2_Bayesian/models/model_M0d.rds")
formula(model_M0d)
fixef(model_M0d)
```

@@

```{r data-priors-M0c}
data_M0c <- ptb_data %>% 
  select(pid, event, period_9) %>%
  mutate(period_9_sq = period_9^2)

priors_M0c <- c(
  set_prior("skew_normal(-1,1,-2)", class = "b", coef = "Intercept"),  
  set_prior("normal(0.1,.1)", class = "b", coef = "period_9"), 
  set_prior("normal(0,.06)", class= "b", coef="period_9_sq"),   
  set_prior("normal(0, 1)", class = "sd"),  
  set_prior("lkj(2)", class = "cor")  
)
```

@@

model_M0c <-
   brm(data = data_M0c,
       family = bernoulli(link="cloglog"),
       formula = event ~ 0 + Intercept + period_9 + period_9_sq + 
              (0 + Intercept + period_9 + period_9_sq | pid),
       prior = priors_M0c,
       chains = 4, cores = 4, iter = 3000, warmup = 1000,
       control = list(adapt_delta = 0.999, step_size = 0.04, max_treedepth = 12),
       seed = 12, init = "0",
       file = "Tutorial_2_Bayesian/models/model_M0c")
```

Model_M0c took about 71 minutes to run.

@@

```{r}
dat <- tibble(x = -3:6,
              x2 = x^2,
              x3 = x^3,
              cloglog1 = -1.8   + 0.3*x   - 0.1*x2   + 0.009*x3,
              cloglog2 = -1     + 0.4*x   - 0.1*x2   + 0.007*x3,
              cloglog3 = -3     - 0.1*x   + 0.12*x2  - 0.002*x3,
              cloglog4 = -0.40  + 0.2*x   - 0.04*x2  + 0.003*x3,
              haz1 = 1 - exp(-1*exp(cloglog1)),
              haz2 = 1 - exp(-1*exp(cloglog2)),
              haz3 = 1 - exp(-1*exp(cloglog3)),
              haz4 = 1 - exp(-1*exp(cloglog4)))

@@

#set_prior("normal(0.1,.08)",class="b",coef="period_9"), # ok !!!
#set_prior("normal(0,.06)",class="b",coef="period_9_sq"),

set_prior("normal(0.1,.1)",class="b",coef="period_9"), # ok !!!
set_prior("normal(0,.06)",class="b",coef="period_9_sq"),
#set_prior("normal(0,.003)",class="b",coef="period_9_cu"),

@@

model_M0_lin_prior454 <-
   brm(data = M0c_data,
       family = binomial(link="cloglog"),
       event | trials(1) ~ 0 + Intercept + period_9 + period_9_sq  +
                          (0 + Intercept + period_9 + period_9_sq  | pid),
       prior = priors_M0c_prior,
       chains = 4, cores = 4, iter = 3000, warmup = 1000,
       control = list(adapt_delta = 0.999, step_size = 0.04, max_treedepth = 12),
       seed = 12, init = "0",
       sample_prior = "only",
       file = "Tutorial_2_Bayesian/models/model_M0_lin_prior454")
```

```{r}
#model_M0_lin_prior448 <- readRDS("Tutorial_2_Bayesian/models/model_M0_lin_prior448.rds")

post <- as_draws_df(model_M0_lin_prior454) %>% 
  select(-lp__) %>% 
  as_tibble()

@@

# plot
p_tidy_fixed <- ggplot(tidy_fixed, aes(x = name, y = value, fill = name)) +  
  stat_halfeye(alpha=0.7, point_interval = "median_qi",.width=c(.8,.99) ) 
p_tidy_fixed

@@

N_obs = 100 # increase for smoother results

# prepare columns of data set
df_pred <- tibble(period_9 = integer(0),
                 period_9_sq = integer(0),
                # period_9_ce = integer(0),
                 pid = integer(0),
                  .row = integer(0),
                  .chain = integer(0),
                  .iteration = integer(0),
                  .draw = integer(0),
                  .prediction = integer(0),
                  obs = integer(0))

@@

for(i in 1:N_obs){
  prior_pred <- add_predicted_draws(model_M0_lin_prior454, 
                                    newdata = newdata_prior, 
                                    summary=F, 
                                    ndraws=NULL) %>% 
                mutate(obs = i)
  df_pred <- bind_rows(df_pred, prior_pred)
}

@@

ggplot(df_pred_haz %>% 
   filter(.draw <639), aes(x=period_9, y=pred_haz, group=.draw)) +
   geom_line(color="black") +
   geom_line(data=df_pred_haz %>% filter(.draw ==sel[1]), 
             color="red", linewidth=2) +
  geom_line(data=df_pred_haz %>% filter(.draw == sel[2]), 
            color="green", linewidth=2) +
  geom_line(data=df_pred_haz %>% filter(.draw == sel[3]), 
          color="blue", linewidth=2) +
  scale_y_continuous(limits = c(0,1)) +
  scale_x_continuous(limits = c(-3,6), breaks = c(-3:6)) +
  facet_wrap(~pid)
```
